---
title: "Supporting information - Fisher's alpha is generally not a valid metric of biodiversity in tropical forests"
# Output formats
format:
  stylisharticle-html:
    # Color example, see "Text color"
    css: colors.css
  elsevier-pdf:
    journal:
      name: Ecological Modelling
      formatting: preprint
      model: 1p
      layout: onecolumn
      cite-style: authoryear
# Bibliography
reference-section-title: References
bibliography: references.bib
csl: global-ecology-and-biogeography
# Language
lang: en-GB
# Code options 
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: true
# Template specific
journalinfo: "Preprint"
archive: "DOI: xxx/xx"
keywordlabel: Keywords
corrauthorlabel: Corresponding Author
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitR", "formatR", "kableExtra"))

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c("tidyverse", "divent", "vegan", "untb", "sads")
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = FALSE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

# Estimating $\alpha$ with R {#sec-appendix1 .appendix}

## Point estimation {.appendix}

Most implementations of the calculation of $\alpha$ are limited to solving the equation

$$S(n) = \alpha \ln \left( 1 + \frac{n}{\alpha} \right),$$

where $S(n)$ is the number of species observed in a sample of size $n$.
The only data required are the total number of species and the sample size.
The fifty hectares of BCI are used in the examples below.

```{r}
library("tidyverse")
library("vegan")
data(BCI)
BCI %>% 
  colSums() ->
  BCI_50ha
```

Functions are:

-   `vegan::fisher.alpha()` that relies on `vegan::fisherfit()`. The numerical resolution uses `uniroot()`.

```{r}
#| echo: true
library("vegan")
fisher.alpha(BCI_50ha)
fisherfit(BCI_50ha)
```

-   `untb::fishers.alpha()`, identical to `vegan::fisherfit()`

```{r}
#| echo: true
library("untb")
fishers.alpha(N = sum(BCI_50ha), S = sum(BCI_50ha > 0))
```

## Likelihood maximisation {.appendix}

The alternative is to adjust the observed abundance distribution to the theoretical distribution of a log series, maximising its likelihood as a function of $\alpha$.

This is done by `sads::fitls()` which uses `mle2()` to maximise the likelihood of `sads::dls()` (the probability density of the log-series).
The default starting value is the point estimate.

```{r}
#| echo: true
library("sads")
fitls(BCI_50ha) %>% 
  coef()
```

In the case of BCI, the result is the same: the likelihood calculation gives predominant weight to the most abundant species.

## Goodness of fit {.appendix}

The theoretical rank of a species $r_s$ can be calculated from its abundance $n_s$ [@May1975] with the following function `lseries_RAC`:

$$r_s = \alpha \int_{n_s \ln(1 + \alpha / n)}^\infty{\frac{e^{-t}}{t} dt}$$

It is a real value, not an integer.

```{r}
#| label: lseries_RAC
#| echo: true

#' Theoretical rank of abundances
#'
#' @param abd a vector of abundances
#' @param alpha Fisher's alpha of the distribution
#'
#' @returns a vector of ranks
lseries_RAC <- function(abd, alpha) {
  n <- sum(abd)
  rank <- vapply(
    # Calculate the rank of each species
    abd, 
    FUN = function(n_s) {
      # Calculate the integral
      f <- stats::integrate(
        function(t) {exp(-t) / t}, 
        lower = n_s * log(1 + alpha / n), 
        upper = Inf
      )
      # Multiply it by alpha
      alpha * f[["value"]]
    }, 
    FUN.VALUE = 0
  )
  # Return the RAC as a tibble
  tibble::tibble(rank, abundance = abd) |>
    dplyr::arrange(rank)
}
```

The best-fit value of $\alpha$ can be found by minimizing the sum of the absolute differences between theoretical and observed ranks.

```{r}
#| label: fit_alpha
#| echo: true

#' Fit alpha by minimizing the departure of species 
#' ranks from their theoretical value in a log-series
#' @param abd a vector of abundances
#'
#' @returns the best-fit value of of Fisher's alpha
fit_alpha <- function(abd) {
  abd_decr <- sort(abd[abd > 0], decreasing = TRUE)
  optimized <- stats::optim(
    par = vegan::fisher.alpha(abd),
    fn = function(alpha) {
      sum(abs(seq_along(abd_decr) 
        - lseries_RAC(abd_decr, alpha)$rank))
    }, 
    method = "L-BFGS-B",
    lower = 0
  )
  optimized$par
}
```

Applied to BCI, the estimate of $\alpha$ is `r round(fit_alpha(BCI_50ha))`.

```{r}
#|label: bci_alpha_gof_code
library("divent")
BCI_50ha %>% 
  as_abundances() %>% 
  autoplot() +
  geom_line(
    data = lseries_RAC(BCI_50ha, alpha = fit_alpha(BCI_50ha)),
    aes(x = rank, y = abundance)    
  ) ->
  fig_hubbell_alpha
```

::: {#suppfig-bci_alpha_gof}
```{r}
#| label: bci_alpha_gof
fig_hubbell_alpha
```

Rank-Abundance Curve (RAC) of BCI (red points) and of the best-fit log-series (black line).
:::

# Estimating $\theta$ by the abundant species distribution {#sec-appendix2 .appendix}

@Hubbell2001 [page 293] proposes estimating $\theta$ by adjusting the rank-abundance curve to the most frequent species in the community, arguing that these species are also the most abundant in the metacommunity.

```{r}
#| label: hubbell_alpha_code
#| echo: true
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  subset(. > median(.)) ->
  BCI_abundant
BCI_theta <- fit_alpha(BCI_abundant)
library("divent")
BCI_abundant %>% 
  as_abundances() %>% 
  autoplot() +
  geom_line(
    data = lseries_RAC(BCI_50ha, alpha = BCI_theta),
    aes(x = rank, y = abundance)    
  ) ->
  fig_hubbell_alpha
```

::: {#suppfig-hubbell_alpha}
```{r}
#| label: hubbell_alpha
fig_hubbell_alpha
```

Estimation of $\theta$ at BCI based on the best fit of $\alpha$, assuming the half most abundant species are distributed as a log-series.
The rank-abundance curve of the data (red points) and of the inferred log-series (black line) are shown.
:::

The estimate using Hubbell's method is `r round(BCI_theta)`.
The good fit of this estimation is shown on @suppfig-hubbell_alpha.

# Full simulation {#sec-appendix3 .appendix}

The objective is to close the loop: using the parameters obtained from the BCI data, simulate a metacommunity and then a community of the same size as BCI.
To validate the approach, the simulated distribution should be similar to the actual one.

BCI parameters $\theta$ and $m$ are estimated following @Etienne2005 with the untb package.

```{r}
#| label: bci_params
#| echo: true
library("untb")
if (file.exists("data/BCI_logkda.RData")) {
  # Use the saved result if available to save
  # computation resources
  load("data/BCI_logkda.RData")
} else {
  BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # requires pari/gp installed and in the search paths,
  # or use gp_binary argument
  logkda() ->
  BCI_logkda
  # Save the result
  save(BCI_logkda, file = "data/BCI_logkda.RData")
}
BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # Estimate theta and m
  optimal.params(log.kda = BCI_logkda) %>% 
  print() ->
  BCI_params
```

A metacommunity with $J_m = 10^6$ (arbitrary) and $\theta$ derived from BCI data is simulated.

```{r}
#| label: the_metacommunity
#| echo: true
J_m <- 1E6
theta <- BCI_params[1]
library("divent")
the_metacommunity <- rcommunity(
  1, 
  size = J_m , 
  distribution = "lseries", 
  fisher_alpha = theta
)
```

The distribution of the local community is derived.
$J$ is that of BCI and $m$ was estimated previously, compared to an arbitrary value of $m$.

```{r}
#| label: bci_sim_code
#| echo: true
J <- sum(BCI)
m <- BCI_params[2]
library("sads")
# Number of species in the local community
Svolkov(theta, m, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances
  rvolkov(theta, m, J) %>% 
  # Make it an abundances object top plot it
  as_abundances() %>% 
  # Whittaker plot
  autoplot() ->
  BCI_sim_neutral
# BCI real distribution
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_50ha
# Compare with m = 1%
m_arbitrary <- .01
Svolkov(theta, m_arbitrary, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances
  rvolkov(theta, m, J) %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_m001
# Plot altogether
BCI_sim_neutral +
  geom_line(
    data = BCI_tbl_50ha,
    aes(x = rank, y = value)
  ) +
  geom_point(
    data = BCI_tbl_m001,
    aes(x = rank, y = value),
    color = "blue"
  ) ->
  fig_bci_sim
```

::: {#suppfig-bci_sim}
```{r}
#| label: bci_sim
fig_bci_sim
```

RAC of BCI compared to two simulated distributions.
The simulated zero-sum multinomial distribution (red dots) simulated with the $\theta$ and $m$ parameters derived from the data corresponds fairly well to the actual distribution of BCI (black line).
The blue dots are a distribution simulated with BCI's $\theta$ but a migration parameter reduced to 0.01 to illustrate that the same values of $\theta$ correspond to very different communities.
Both $\theta$ and $m$ must be considered to describe the diversity.
:::

@suppfig-bci_sim shows that the simulated community is very similar to the real one but that modifying the migration parameter results in a very different distribution: $\theta$ alone does not give much information on the local community.

# Increasing regional sampling {#sec-appendix4 .appendix}

An increasing number of GuyaDiv plots (one per site) is randomly selected to sample the metacommunity.
The estimated $\alpha$ is plotted against the number of plots (@suppfig-guyadiv_alpha).
It increases with the logarithm of the number of plots that is a proxy of the size of the sampled metacommunity.

```{r}
#| label: guyadiv_alpha_code
#| echo: true
# Read the data
# Plots
"data/plots.csv" %>% 
  read_csv2() ->
  guyadiv_plots
# Abundances
"data/abundances.csv" %>% 
  read_csv2() ->
  guyadiv_abd
# Number of simulations
simulations_n <- 20
# Function to randomly select a plot in each of a
# chosen number of sites (sample_size), 
# merge the plots and return alpha.
guyadiv_sub_alpha <- function(sample_size) {
  # Select one plot per location...
  guyadiv_plots %>% 
    # Set a random value to each plot
    mutate(random = runif(n())) %>% 
    # Save the tibble
    {. ->> plots_randomized} %>% 
    # Select the plot with the max random value 
    # in each location
    group_by(Location) %>%
    summarize(random_max = max(random)) %>% 
    rename(random = random_max) %>% 
    # Eliminate non-selected plots
    inner_join(plots_randomized) %>% 
    # Keep the plot number and the random number
    dplyr::select(Plot, random) %>% 
    # Get the abundances
    inner_join(guyadiv_abd) %>% 
    # Sort by random number
    arrange(random) %>% 
    # Keep the first rows
    head(sample_size) %>% 
    # Delete the random numbers
    dplyr::select(-random) %>% 
    # Make the table an abundance object (divent)
    rename(site = Plot) %>% 
    as_abundances() %>% 
    # Give all plots equal weight
    mutate(weight = 1) %>% 
    # Merge the inventories
    metacommunity() %>% 
    # Extract a numeric vector
    as.numeric() %>% 
    # Compute alpha
    fisher.alpha()
}

# Count the number of sites
guyadiv_plots %>% 
  group_by(Location) %>% 
  summarize() %>% 
  nrow() ->
  sites_n
# Choose sizes between 3 and the number of sites
sizes <- seq_len(sites_n)[-(1:2)]
# Calculate alpha of each sample
alphas <- replicate(
  simulations_n,
  expr = sapply(sizes, FUN = guyadiv_sub_alpha)
)
tibble(NbPlots = sizes) %>%
  bind_cols(alphas) %>% 
  pivot_longer(cols = !NbPlots) %>% 
  ggplot(aes(x = NbPlots, y = value)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Plots", y = "Fisher's alpha") +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2)) ->
  fig_guyadiv_alpha
```

::: {#suppfig-guyadiv_alpha}
```{r}
#| label: guyadiv_alpha
fig_guyadiv_alpha
```

Estimation of the metacommunity's $\alpha$ as a function of the number of GuyaDiv plots used to sample it.
The line is the fit of the quadratic model relating $\alpha$ to the logarithm of the number of plots.
:::

\pagebreak

# Code to reproduce the main text

## Figure 1

```{r}
#| label: BCI_accum_logx
#| echo: true
library("tidyverse")
library("divent")
library("vegan")
data(BCI)
BCI %>% 
  # Sum all plots
  colSums() %>% 
  # Save the vector of abundances
  {. ->> BCI_50ha} %>% 
  # Calculate alpha
  fisher.alpha() ->
  BCI_alpha
 BCI_50ha %>%
  # Declare the object class (divent)
  as_abundances() %>% 
  # Site name
  mutate(site = "BCI") %>% 
  # Accumulation curve up to 2x sample size
  accum_hill(levels = seq_len(2 * sum(BCI_50ha))) %>% 
  # Save the SAC
  {. ->> BCI_accum} %>% 
  # Plot
  autoplot() +
  scale_x_log10() +
  theme(legend.position = "none") +
  labs(y = "Number of Species") +
  # Accumulation curve of a log-series
  geom_function(fun = \(n) {BCI_alpha * log(1 + n / BCI_alpha)}, col = "black")
```

Species Accumulation Curve (SAC) of a log-series distribution (black) with Fisher's $\alpha$ equal to `r round(BCI_alpha)` and of BCI (red).
The intersection of the dotted lines corresponds to the number of trees and of species of BCI 50-ha inventory the value of $\alpha$ is derived from.

## Figure 2

```{r}
#| label: BCI_RAC
#| echo: true
BCI_50ha %>%
  fit_rac(distribution = "lseries") %>% 
  pluck("rac") ->
  BCI_ls_RAC
BCI_50ha %>%
  # Declare the object class (divent)
  as_abundances() %>% 
  autoplot() + 
  geom_line(
    data  = BCI_ls_RAC,
    aes(x = rank, y = abundance)
  )
```

Rank-Abundance Curve (RAC) of a log-series distribution (black) and of BCI (red dots).
Species are ordered by decreasing abundance.
The log-series is the best fit to BCI data ($\alpha =$ `r round(BCI_alpha)`).

## Figure 3

```{r}
#| label: BCI_accum
#| echo: true
BCI_S_estimate <- div_richness(BCI_50ha, as_numeric = TRUE)
BCI_density <- round(sum(BCI) / 50)
BCI_S_1ha <- round(
  div_richness(
    BCI_50ha, 
    level = BCI_density, 
    as_numeric = TRUE
  )
)
BCI_accum %>%
  autoplot() +
  labs(y = "Richness") +
  theme(legend.position = "none") +
  # Asymptotic estimator
  geom_hline(
    yintercept = BCI_S_estimate, 
    linetype = 3
  ) +
  # Accumulation curve of a log-series
  geom_function(fun = \(n) {BCI_alpha * log(1 + n / BCI_alpha)}, col = "black")
```

Species Accumulation Curve of BCI extrapolated up to twice the sample size.
The dotted lines correspond to the 50-ha inventory.
The horizontal, dotted line, is the estimated asymptotic number of species, calculated with the jackknife estimator of order 1.
The black curve is the SAC of the log-series with $\alpha$ equal to `r round(BCI_alpha)`.

## Figure 4

```{r}
#| label: BCI_size
# Sample `sample_size` sites in an abundance table, group them and compute alpha
alpha_sub <- function(sample_size, abd) {
  # Draw the rows of to sample in the table
  the_sample <- sample(seq_len(nrow(abd)), size = sample_size) 
  # Sample them
  abd[the_sample, ] %>% 
    # Make an abundances object (package divent)
    as_abundances() %>% 
    # Equal weights
    mutate(weight = 1) %>% 
    # Merge the rows
    metacommunity() %>% 
    # Numeric vector
    as.numeric() %>% 
    # Return alpha
    fisher.alpha()
}

# Sample from 4 to 50ha
sizes <- seq_len(nrow(BCI))[-(1:3)]
# Number of simulations
simulations_n <- 20
# Compute alpha
alphas <- replicate(
  simulations_n, 
  # Caclulate alpha at all sample sizes
  expr = sapply(sizes, FUN = alpha_sub, abd = BCI)
)
# alpha values
tibble(Size = sizes) %>%
  bind_cols(alphas) %>% 
  pivot_longer(cols = !Size) %>% 
  # Plot
  ggplot(aes(x = Size, y = value)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Sample size", y = "Fisher's alpha") +
  geom_smooth(method = "lm")
```

Estimation of Fisher's $\alpha$ depending on the number of 1-ha plots of BCI.
The dots are random draws of plot assemblages.
The line is the fit of a linear model relating $\alpha$ to the logarithm of the number of plots.

## Figure 5

The `plot.csv` file was published in @Marcon2024 and is available on [Github](https://github.com/EricMarcon/JTE-22-105/blob/master/data/plots.csv).

```{r}
#| label: arrhenius_logx
# Read the data
# Plots
"data/plots.csv" %>% 
  read_csv2() ->
  guyadiv_plots
# Abundances
"data/abundances.csv" %>% 
  read_csv2() ->
  guyadiv_abd
# Estimate alpha
guyadiv_abd %>% 
  # Make an abundances object (package divent)
  rename(site = Plot) %>% 
  as_abundances() %>% 
  # Equal weights
  mutate(weight = 1) %>% 
  # Merge the rows
  metacommunity() %>% 
  # Numeric vector
  as.numeric() %>% 
  # Return alpha
  fisher.alpha() ->
  gf_alpha_guyadiv
# Marcon et al. (2024) results
c <- 694
z <- .104
gf_density <- 627 * 100
gf_area <- 8E4
# Number of species according to Guyadiv alpha
gf_ls_species <- gf_alpha_guyadiv * log(1 + gf_density * gf_area / gf_alpha_guyadiv)
# alpha according to the power model
library("untb")
gf_alpha_arrhenius <- fishers.alpha(
  N = gf_density * gf_area, 
  S = c * gf_area^z
)
ggplot(data.frame(x = 10^(1:6)), aes(x)) +
  geom_function(fun = \(A) c * A^z) +
  scale_x_log10() +
  geom_vline(
    xintercept = gf_area, 
    linetype = 2
  ) +
  geom_hline(
    yintercept = c * gf_area^z, 
    linetype = 2
  ) +
  # Accumulation curve of log-series
  geom_function(
    fun = \(A) {gf_alpha_guyadiv * log(1 + A * gf_density / gf_alpha_guyadiv)}, 
    col = "blue"
  ) +
  geom_function(
    fun = \(A) {gf_alpha_arrhenius * log(1 + A * gf_density / gf_alpha_arrhenius)}, 
    col = "red"
  ) +
  labs(x = "Area", y = "Number of species")
```

Number of tree species depending on the area (in square km, with density `r gf_density/100` trees/ha) in French Guiana, according to the Arrhenius's power model estimated by @Marcon2024.
The Y-axis is not in logarithm, so the relation is a convex curve instead of a straight line.
The black dotted lines show the valid estimation of richness in the whole territory: around 2200 species for 80,000 km^2^.
The blue line is the species accumulation curve of the log-series estimated from the GuyaDiv metacommunity assuming its area is that of French Guiana.
It overestimates the regional biodiversity at all scales.
The red curve is the SAC of the actual metacommunity of French Guiana, whose Fisher's $\alpha$ is twice smaller.
