---
title: "Appendix - Fisher's alpha is generally not a valid metric of biodiversity in tropical forests"
# Output formats
format:
  stylisharticle-html:
    # Color example, see "Text color"
    css: colors.css
  elsevier-pdf:
    journal:
      name: Ecological Modelling
      formatting: preprint
      model: 1p
      layout: onecolumn
      cite-style: authoryear
author:
  - name: Eric Marcon
    affiliations:
      - name: AgroParisTech
        department: UMR Amap, Université de Montpellier, Cirad, CNRS, INRAE, IRD
        city: Montpellier
        country: France
    orcid: 0000-0002-5249-321X
    email: eric.marcon@agroparistech.fr
    url: https://ericmarcon.github.io/
    attributes:
      corresponding: true
# Abstract and Keywords
abstract: |
  Fisher's alpha should not be used to assess the diversity of tropical forest tree species at the local and regional scales.
  At the local scale, communities are not distributed in log-series: alpha is the parameter of a model that does not fit the data so it can neither be estimated nor interpreted unambiguously.
  At the regional scale, alpha correctly describes the distribution of a metacommunity whose extent is unknown when it is not the whole forest area.
  Species richness is always a better diversity metric than alpha, containing basically the same information without relying on uncertain assumptions.
keywords: [Fisher's alpha, biodiversity, tropical forests]
# Bibliography
reference-section-title: References
bibliography: references.bib
csl: elsevier-harvard
# Language
lang: en-GB
# Code options 
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: true
# Template specific
journalinfo: "Preprint"
archive: "DOI: xxx/xx"
keywordlabel: Keywords
corrauthorlabel: Corresponding Author
crossref:
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      space-before-numbering: false
      latex-list-of-description: Supplementary Figure
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitR", "formatR", "kableExtra"))

# Chunk font size hook: allows size='small' 
# or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c("tidyverse", "divent", "vegan", "untb", "sads")
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = FALSE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```


# Estimating $\alpha$ with R {#sec-appendix1 .appendix}

## Point estimation {.appendix}

Most implementations of the calculation of $\alpha$ are limited to solving the equation

$$S(n) = \alpha \ln \left( 1 + \frac{n}{\alpha} \right),$$

where $S(n)$ is the number of species observed in a sample of size $n$.
The only data required are the total number of species and the sample size.
The fifty hectares of BCI are used in the examples below.

```{r}
library("tidyverse")
library("vegan")
data(BCI)
BCI %>% 
  colSums() ->
  BCI_50ha
```

Functions are:

-   `vegan::fisher.alpha()` that relies on `vegan::fisherfit()`. The numerical resolution uses `uniroot()`.

```{r}
#| echo: true
library("vegan")
fisher.alpha(BCI_50ha)
fisherfit(BCI_50ha)
```

-   `untb::fishers.alpha()`, identical to `vegan::fisherfit()`

```{r}
#| echo: true
library("untb")
fishers.alpha(N = sum(BCI_50ha), S = sum(BCI_50ha > 0))
```

## Likelihood maximisation {.appendix}

The alternative is to adjust the observed abundance distribution to the theoretical distribution of a log series, maximising its likelihood as a function of $\alpha$.

This is done by `sads::fitls()` which uses `mle2()` to maximise the likelihood of `sads::dls()` (the probability density of the log-series).
The default starting value is the point estimate.

```{r}
#| echo: true
library("sads")
fitls(BCI_50ha) %>% 
  coef()
```

In the case of BCI, the result is the same: the likelihood calculation gives predominant weight to the most abundant species.

## Goodness of fit {.appendix}

The theoretical rank of a species $r_s$ can be calculated from its abundance $n_s$ [@May1975] with the following function `lseries_RAC`:

$$r_s = \alpha \int_{n_s \ln(1 + \alpha / n)}^\infty{\frac{e^{-t}}{t} dt}$$

It is a real value, not an integer.

```{r}
#| label: lseries_RAC
#| echo: true

#' Theoretical rank of abundances
#'
#' @param abd a vector of abundances
#' @param alpha Fisher's alpha of the distribution
#'
#' @returns a vector of ranks
lseries_RAC <- function(abd, alpha) {
  n <- sum(abd)
  rank <- vapply(
    # Calculate the rank of each species
    abd, 
    FUN = function(n_s) {
      # Calculate the integral
      f <- stats::integrate(
        function(t) {exp(-t) / t}, 
        lower = n_s * log(1 + alpha / n), 
        upper = Inf
      )
      # Multiply it by alpha
      alpha * f[["value"]]
    }, 
    FUN.VALUE = 0
  )
  # Return the RAC as a tibble
  tibble::tibble(rank, abundance = abd) |>
    dplyr::arrange(rank)
}
```

The best-fit value of $\alpha$ can be found by minimizing the sum of the absolute differences between theoretical and observed ranks.

```{r}
#| label: fit_alpha
#| echo: true

#' Fit alpha by minimizing the departure of species 
#' ranks from their theoretical value in a log-series
#' @param abd a vector of abundances
#'
#' @returns the best-fit value of of Fisher's alpha
fit_alpha <- function(abd) {
  abd_decr <- sort(abd[abd > 0], decreasing = TRUE)
  optimized <- stats::optim(
    par = vegan::fisher.alpha(abd),
    fn = function(alpha) {
      sum(abs(seq_along(abd_decr) 
        - lseries_RAC(abd_decr, alpha)$rank))
    }, 
    method = "L-BFGS-B",
    lower = 0
  )
  optimized$par
}
```

Applied to BCI, the estimate of $\alpha$ is `r round(fit_alpha(BCI_50ha))`.

```{r}
#|label: bci_alpha_gof_code
library("divent")
BCI_50ha %>% 
  as_abundances() %>% 
  autoplot() +
  geom_line(
    data = lseries_RAC(BCI_50ha, alpha = fit_alpha(BCI_50ha)),
    aes(x = rank, y = abundance)    
  ) ->
  fig_hubbell_alpha
```

::: {#suppfig-bci_alpha_gof}
```{r}
#| label: bci_alpha_gof
fig_hubbell_alpha
```

Rank-Abundance Curve (RAC) of BCI (red points) and of the best-fit log-series (black line).
:::

# Estimating $\theta$ by the abundant species distribution {#sec-appendix2 .appendix}

@Hubbell2001 [page 293] proposes estimating $\theta$ by adjusting the rank-abundance curve to the most frequent species in the community, arguing that these species are also the most abundant in the metacommunity.

```{r}
#| label: hubbell_alpha_code
#| echo: true
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  subset(. > median(.)) ->
  BCI_abundant
BCI_theta <- fit_alpha(BCI_abundant)
library("divent")
BCI_abundant %>% 
  as_abundances() %>% 
  autoplot() +
  geom_line(
    data = lseries_RAC(BCI_50ha, alpha = BCI_theta),
    aes(x = rank, y = abundance)    
  ) ->
  fig_hubbell_alpha
```

::: {#suppfig-hubbell_alpha}
```{r}
#| label: hubbell_alpha
fig_hubbell_alpha
```

Estimation of $\theta$ at BCI based on the best fit of $\alpha$, assuming the half most abundant species are distributed as a log-series.
The rank-abundance curve of the data (red points) and of the inferred log-series (black line) are shown.
:::

The estimate using Hubbell's method is `r round(BCI_theta)`.
The good fit of this estimation is shown on @suppfig-hubbell_alpha.

# Full simulation {#sec-appendix3 .appendix}

The objective is to close the loop: using the parameters obtained from the BCI data, simulate a metacommunity and then a community of the same size as BCI.
To validate the approach, the simulated distribution should be similar to the actual one.

BCI parameters $\theta$ and $m$ are estimated following @Etienne2005 with the untb package.

```{r}
#| label: bci_params
#| echo: true
library("untb")
if (file.exists("data/BCI_logkda.RData")) {
  # Use the saved result if available to save
  # computation resources
  load("data/BCI_logkda.RData")
} else {
  BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # requires pari/gp installed and in the search paths,
  # or use gp_binary argument
  logkda() ->
  BCI_logkda
  # Save the result
  save(BCI_logkda, file = "data/BCI_logkda.RData")
}
BCI_50ha %>% 
  # name conflict with dplyr::count()
  untb::count() %>% 
  # Estimate theta and m
  optimal.params(log.kda = BCI_logkda) %>% 
  print() ->
  BCI_params
```

A metacommunity with $J_m = 10^6$ (arbitrary) and $\theta$ derived from BCI data is simulated.

```{r}
#| label: the_metacommunity
#| echo: true
J_m <- 1E6
theta <- BCI_params[1]
library("divent")
the_metacommunity <- rcommunity(
  1, 
  size = J_m , 
  distribution = "lseries", 
  fisher_alpha = theta
)
```

The distribution of the local community is derived.
$J$ is that of BCI and $m$ was estimated previously, compared to an arbitrary value of $m$.

```{r}
#| label: bci_sim_code
#| echo: true
J <- sum(BCI)
m <- BCI_params[2]
library("sads")
# Number of species in the local community
Svolkov(theta, m, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances
  rvolkov(theta, m, J) %>% 
  # Make it an abundances object top plot it
  as_abundances() %>% 
  # Whittaker plot
  autoplot() ->
  BCI_sim_neutral
# BCI real distribution
BCI_50ha %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_50ha
# Compare with m = 1%
m_arbitrary <- .01
Svolkov(theta, m_arbitrary, J) %>% 
  # Must be integers for the simulation
  round() %>% 
  # Draw the species abundances
  rvolkov(theta, m, J) %>% 
  sort(decreasing = TRUE) %>% 
  as_tibble() %>% 
  mutate(rank = seq_len(n())) -> 
  BCI_tbl_m001
# Plot altogether
BCI_sim_neutral +
  geom_line(
    data = BCI_tbl_50ha,
    aes(x = rank, y = value)
  ) +
  geom_point(
    data = BCI_tbl_m001,
    aes(x = rank, y = value),
    color = "blue"
  ) ->
  fig_bci_sim
```

::: {#suppfig-bci_sim}
```{r}
#| label: bci_sim
fig_bci_sim
```

RAC of BCI compared to two simulated distributions.
The simulated zero-sum multinomial distribution (red dots) simulated with the $\theta$ and $m$ parameters derived from the data corresponds fairly well to the actual distribution of BCI (black line).
The blue dots are a distribution simulated with BCI's $\theta$ but a migration parameter reduced to 0.01 to illustrate that the same values of $\theta$ correspond to very different communities.
Both $\theta$ and $m$ must be considered to describe the diversity.
:::

@suppfig-bci_sim shows that the simulated community is very similar to the real one but that modifying the migration parameter results in a very different distribution: $\theta$ alone does not give much information on the local community.

# Increasing regional sampling {#sec-appendix4 .appendix}

An increasing number of GuyaDiv plots (one per site) is randomly selected to sample the metacommunity.
The estimated $\alpha$ is plotted against the number of plots (@suppfig-guyadiv_alpha).
It increases with the logarithm of the number of plots that is a proxy of the size of the sampled metacommunity.

```{r}
#| label: guyadiv_alpha_code
#| echo: true
# Read the data
# Plots
"data/plots.csv" %>% 
  read_csv2() ->
  guyadiv_plots
# Abundances
"data/abundances.csv" %>% 
  read_csv2() ->
  guyadiv_abd
# Number of simulations
simulations_n <- 20
# Function to randomly select a plot in each of a
# chosen number of sites (sample_size), 
# merge the plots and return alpha.
guyadiv_sub_alpha <- function(sample_size) {
  # Select one plot per location...
  guyadiv_plots %>% 
    # Set a random value to each plot
    mutate(random = runif(n())) %>% 
    # Save the tibble
    {. ->> plots_randomized} %>% 
    # Select the plot with the max random value 
    # in each location
    group_by(Location) %>%
    summarize(random_max = max(random)) %>% 
    rename(random = random_max) %>% 
    # Eliminate non-selected plots
    inner_join(plots_randomized) %>% 
    # Keep the plot number and the random number
    dplyr::select(Plot, random) %>% 
    # Get the abundances
    inner_join(guyadiv_abd) %>% 
    # Sort by random number
    arrange(random) %>% 
    # Keep the first rows
    head(sample_size) %>% 
    # Delete the random numbers
    dplyr::select(-random) %>% 
    # Make the table an abundance object (divent)
    rename(site = Plot) %>% 
    as_abundances() %>% 
    # Give all plots equal weight
    mutate(weight = 1) %>% 
    # Merge the inventories
    metacommunity() %>% 
    # Extract a numeric vector
    as.numeric() %>% 
    # Compute alpha
    fisher.alpha()
}

# Count the number of sites
guyadiv_plots %>% 
  group_by(Location) %>% 
  summarize() %>% 
  nrow() ->
  sites_n
# Choose sizes between 3 and the number of sites
sizes <- seq_len(sites_n)[-(1:2)]
# Calculate alpha of each sample
alphas <- replicate(
  simulations_n,
  expr = sapply(sizes, FUN = guyadiv_sub_alpha)
)
tibble(NbPlots = sizes) %>%
  bind_cols(alphas) %>% 
  pivot_longer(cols = !NbPlots) %>% 
  ggplot(aes(x = NbPlots, y = value)) +
  geom_point() +
  scale_x_log10() +
  labs(x = "Number of Plots", y = "Fisher's alpha") +
  stat_smooth(method = "lm", formula = y ~ poly(x, 2)) ->
  fig_guyadiv_alpha
```

::: {#suppfig-guyadiv_alpha}
```{r}
#| label: guyadiv_alpha
fig_guyadiv_alpha
```

Estimation of the metacommunity's $\alpha$ as a function of the number of GuyaDiv plots used to sample it.
The line is the fit of the quadratic model relating $\alpha$ to the logarithm of the number of plots.
:::
